---
title: "Montreal real estate stats"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    self_contained: False
    dev: svg
---
```{r setup,include=FALSE,eval=TRUE, echo=FALSE, message=FALSE,warning=FALSE,cache=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE)
#knitr::opts_chunk$set(warning=FALSE, message=FALSE)
library(flexdashboard)
library(tidyverse)
library(sf)
library(tmap)
library(tmaptools)
library(leaflet)
library(raster)
library(mgcv)
# for survival analysis
library(survival)
library(survminer)

# working with time
library(lubridate)
theme_set(theme_bw(base_size = 10, base_family = "Arial")+
  theme(
        axis.text   = element_text(vjust = 0.2, size = 10),
        axis.title = element_text(face = 'bold', vjust = 0.2, size = 10),
        plot.title = element_text(face = 'bold', vjust = 2.0, size = 12),
        strip.text = element_text(face = 'bold', size = 10),
        plot.margin = unit(c(1.0,0.2,0.2,0.2), "cm"),
        legend.position='bottom'
    ))

```
```{r load_data,eval=TRUE, echo=FALSE, message=FALSE,warning=FALSE,cache=TRUE}
load('preprocessed.RData')

```
```{r prepare,eval=TRUE, echo=FALSE, message=FALSE,warning=FALSE}
mtl_arr<-mtl_pa

kijiji_roi<-kijiji_geo_p%>%dplyr::filter(!is.na(bedrooms))%>%st_join(mtl_land,left=F)

rent<-kijiji_roi %>% mutate(bedrooms=as.factor(bedrooms))
rent_<-bind_cols( as.data.frame(rent), as.data.frame(st_coordinates(rent)))%>%rename(x=X,y=Y)

# filter out empty regions
rent_n<-rent_%>%group_by(nom_qr)%>%summarize(n=n())%>%filter(n>50)

mask_rent<-mtl_p%>%inner_join(rent_n,by=c('nom_qr'))%>%st_intersection(mtl_land)

prop_geo_p <- prop_geo_p %>%
  dplyr::filter( type != 'House', mprice>0, area_interior>300 , !is.na(bedrooms) ) %>%
  mutate(bedrooms=droplevels(bedrooms), type=droplevels(type),
         parking=factor(parking>0,levels=c(F,T),labels=c('No parking','With parking')),
         nom_arr=as.factor(nom_arr), nom_qr=as.factor(nom_qr)) %>%
  st_intersection(mtl_land) 

prop_geo_p_lim<-prop_geo_p %>% as.data.frame() %>% group_by(nom_qr,type)%>%
  summarize(
    price_low = quantile(mprice,0.01), price_high = quantile(mprice,0.99),
    area_low = quantile(area_interior,0.01), area_high = quantile(area_interior,0.99),
  ) %>% dplyr::select(nom_qr,type,price_low,price_high,area_low,area_high)


prop_geo_p<- prop_geo_p %>% left_join(prop_geo_p_lim, by=c('nom_qr','type')) %>%
  filter(mprice<=price_high, mprice>=price_low,
         area_interior<=area_high,area_interior>=area_low) %>% 
  dplyr::select(-price_high,-price_low,-area_high,-area_low) 

prop_geo_pp<-bind_cols( as.data.frame(prop_geo_p), as.data.frame(st_coordinates(prop_geo_p)))%>%rename(x=X,y=Y)

condo_geo_pp<-prop_geo_pp%>%filter(type=='Apartment')
plex_geo_pp<-prop_geo_pp%>%filter(type!='Apartment')

condo_geo_ppn<-condo_geo_pp%>%group_by(nom_qr)%>%summarize(n=n())%>%filter(n>5)%>%as.data.frame()
plex_geo_ppn<-plex_geo_pp%>%group_by(nom_qr)%>%summarize(n=n())%>%filter(n>5)%>%as.data.frame()

mask_condo<-mtl_p%>%inner_join(condo_geo_ppn,by=c('nom_qr'))%>%st_intersection(mtl_land)
mask_plex<-mtl_p%>%inner_join(plex_geo_ppn, by=c('nom_qr'))%>%st_intersection(mtl_land)

subway_p<-subway_p%>%mutate(route_name=droplevels(route_name))
```
Using [generalized additive models](https://en.wikipedia.org/wiki/Generalized_additive_model) 
to visualize prices on the Montreal Island. Based on data from kijiji and centris. Gray areas 
mean that not enough reliable information was available.

# Rent 4 1/2

## {data-height=100}

### Time period

```{r rent_ts}
valueBox(paste(min(rent_$first_ts),max(rent_$first_ts)))
```


### N

```{r rent_N}
valueBox(length(rent$ts))
```

## {data-height=900}

### {data-width=700}

```{r rent_map,eval=TRUE, echo=FALSE, message=FALSE,warning=FALSE}
fam=inverse.gaussian(link="log")
ilink <- fam$linkinv

# create spatical model with smooth price variable
model_rent_geo<-gam(price~bedrooms + s(x,y,k=200),
        data=rent_,bs="cr",method='REML',
        family=fam)


pred_rent <- raster(extent(mtl_land),res=100)
crs(pred_rent)<-crs(mtl_land)

# need to predict response (link by default)
my_predict<-function(...) predict(...,type="response")

# predict 4 1/2 rents
pred_rent <- raster::interpolate(pred_rent,model_rent_geo,fun=my_predict, xyOnly=T,
                                const=data.frame(bedrooms=2))
pred_rent <- mask(pred_rent, mask_rent)

m<-tm_shape(mtl_arr,bbox=extent(mtl_arr)) + tm_borders(alpha=0.8, col='black')+
   tm_shape(pred_rent)+tm_raster(style="cont",alpha=0.7, title='$')+
   tm_shape(subway_stop_p%>%dplyr::select(stop_name))+tm_symbols(col='blue',alpha=0.2,size=0.03)+
   tm_shape(subway_p)+tm_lines(col='blue',alpha=0.2)+
   tm_scale_bar(position=c("right", "bottom"))+
   tm_layout(scale=1)

# convert to interactive leaflet map
tmap_leaflet(m)
#m
```

### {data-width=300}
```{r rent_hist}
ggplot(rent,aes(x=bedrooms,y=price))+geom_violin()+
  scale_y_continuous(labels = scales::dollar)+
  ylab('rent/month')+xlab('bedrooms')
```


# Condo 4 1/2 price/sqft

## {data-height=100}

### Time period

```{r condo_ts}
valueBox(paste(min(condo_geo_pp$first_ts),max(condo_geo_pp$first_ts)))
```


### N

```{r condo_N}
valueBox(length(condo_geo_pp$first_ts))
```

## {data-height=900}

### {data-width=700}

```{r condo_map}
model_psqf_geo<-gam(price_sqft ~ bedrooms + s(x,y, k=100),
        data=condo_geo_pp, bs="cr",method='REML',
        family=fam)


pred_price <- raster(extent(mtl_land),res=100)
crs(pred_price)<-crs(mtl_land)

# predict 4 1/2 cost
pred_price <- raster::interpolate(pred_price,model_psqf_geo,fun=my_predict, xyOnly=T,
                                  const=data.frame(bedrooms=2))
pred_price <- mask(pred_price, mask_condo)

m<-tm_shape(mtl_arr,bbox=extent(mtl_arr)) + tm_borders(alpha=0.8, col='black')+
   tm_shape(pred_price)+tm_raster(style="cont",alpha=0.7, title='$/sqft')+
   tm_shape(subway_stop_p%>%dplyr::select(stop_name))+tm_symbols(col='blue',alpha=0.2,size=0.03)+
   tm_shape(subway_p%>%dplyr::select(route_name))+tm_lines(col='blue',alpha=0.2)+
   tm_scale_bar(position=c("right", "bottom"))+
   tm_layout(scale=1)

# convert to interactive leaflet map
tmap_leaflet(m)
```

### {data-width=300} 

```{r condo_hist}
ggplot(condo_geo_pp,aes(x=bedrooms,y=mprice))+geom_violin()+
  scale_y_continuous(labels = scales::dollar)+
  ylab('total price')+xlab('bedrooms')
```


# Duplex with 4 1/2 price/sqft 

## {data-height=100}

### Time period

```{r plex_ts}
valueBox(paste(min(plex_geo_pp$first_ts),max(plex_geo_pp$first_ts)))
```


### N

```{r plex_N}
valueBox(length(plex_geo_pp$first_ts))
```

## {data-height=900}

### {data-width=700}

```{r plex_map}
model_psqf_geo2<-gam(price_sqft ~ type + bedrooms + s(x,y, k=100),
        data=plex_geo_pp, bs="cr",method='REML',
        family=fam)


pred_price2 <- raster(extent(mtl_land),res=100)
crs(pred_price2)<-crs(mtl_land)

# predict 4 1/2 prices
pred_price2 <- raster::interpolate(pred_price2,model_psqf_geo2,fun=my_predict, xyOnly=T,
                                  const=data.frame(type='Duplex', bedrooms=2 ))
pred_price2 <- mask(pred_price2, mask_plex)

m<-tm_shape(mtl_arr,bbox=extent(mtl_arr)) + tm_borders(alpha=0.8, col='black')+
   tm_shape(pred_price2)+tm_raster(style="cont",alpha=0.7, title='$/sqft')+
   tm_shape(subway_stop_p%>%dplyr::select(stop_name))+tm_symbols(col='blue',alpha=0.2,size=0.03)+
   tm_shape(subway_p%>%dplyr::select(route_name))+tm_lines(col='blue',alpha=0.2)+
   tm_scale_bar(position=c("right", "bottom"))+
   tm_layout(scale=1)

# convert to interactive leaflet map
tmap_leaflet(m)
```

### {data-width=300} 

```{r plex_hist}
ggplot(plex_geo_pp,aes(x=bedrooms,y=mprice))+geom_violin()+
  scale_y_continuous(labels = scales::dollar)+
  facet_grid(type~.)+
  ylab('total price')+xlab('bedrooms')
```

# Time course for the whole island

##

###

```{r mtl_timecourse}
start_ts=min(prop_geo_pp$first_ts)
end_ts=max(prop_geo_pp$first_ts)

model_psqft_mtl_t <- gam(price_sqft ~ type +
  s(start_date, by=type, k=6),
          data=prop_geo_pp, bs="cr",
          family=fam, method='REML')


simul_mtl<-expand.grid(type=levels(prop_geo_pp$type),
                   first_ts=seq(start_ts,end_ts,by='day')) %>%
  mutate(start_date=as.numeric(first_ts))

# predict price
ppp<-predict(model_psqft_mtl_t, newdata=simul_mtl, se.fit=T)
simul_mtl$price_sqft<-ilink(ppp$fit)
simul_mtl$price_sqft_upr<-ilink(ppp$fit+2*ppp$se.fit)
simul_mtl$price_sqft_lwr<-ilink(ppp$fit-2*ppp$se.fit)

ggplot( simul_mtl,
       aes(x=first_ts, y=price_sqft, ymin=price_sqft_lwr, ymax=price_sqft_upr, col=type))+
  ylab('$/sqft')+xlab('')+
  scale_y_continuous(labels = scales::dollar)+
  geom_line()+geom_ribbon(alpha=0.4,lty=2)+
  scale_x_date(limits=c(start_ts,end_ts))+
  theme(
        axis.text.x=element_text(angle=60, hjust=1, size=12)
    ) 

```

# Time course by borough

## 

### 

```{r borough_timecourse}
# determine start and end date for each property type for each area

# remove underrepresented ranges
prop_geo_lim<-prop_geo_pp %>% 
  group_by(nom_arr, type) %>%
  summarize( begin_date=min(start_date), 
               end_date=max(start_date))

model_psqft_arr_t <- gam(price_sqft ~ type + nom_arr +  
  s(start_date, by=type, k=4) + s(start_date, by=nom_arr, k=4),
          data=prop_geo_pp, bs="cr",
          family=fam, method='REML')

#model_psqft_arr_t <- gam(price_sqft ~ type + nom_arr + parking + s(start_date, by=type, k=8) ,
#          data=prop_geo_pp, bs="cr",
#          family=fam, method='REML')

#print(summary(model_psqft_arr_t))

# calculate regression values
simul2<-expand.grid(type=levels(prop_geo_pp$type),
                   parking=c('With parking'),
                   bedrooms=c(2),
                   first_ts=seq(start_ts,end_ts,by='day'),
                   nom_arr=levels(prop_geo_pp$nom_arr)) %>%
  mutate(start_date=as.numeric(first_ts))

# define domain 
simul2 <- inner_join(simul2, prop_geo_lim, by=c('type', 'nom_arr') ) %>% 
  filter(start_date>=begin_date,start_date<=end_date)%>%
  mutate(nom_arr=droplevels(nom_arr))

# predict price
ppp<-predict(model_psqft_arr_t, newdata=simul2, se.fit=T)
simul2$price_sqft<-ilink(ppp$fit)
simul2$price_sqft_upr<-ilink(ppp$fit+2*ppp$se.fit)
simul2$price_sqft_lwr<-ilink(ppp$fit-2*ppp$se.fit)

for(n in levels(simul2$nom_arr)) {
gg<-ggplot( simul2%>%filter(nom_arr==n),
       aes(x=first_ts, y=price_sqft, ymin=price_sqft_lwr, ymax=price_sqft_upr, col=type))+
  ylab('')+xlab('')+
  geom_line()+geom_ribbon(alpha=0.4,lty=2)+
  scale_y_continuous(labels = scales::dollar)+
  scale_x_date(limits=c(start_ts,end_ts))+
  theme(
        axis.text.x=element_text(angle=60, hjust=1, size=12)
    ) + ggtitle(n)

print(gg)
}
```


# Spatial survival on the market

##

### {data-width=700}
```{r surv_map}
mmt<-gam(time_on_market ~ type + s(x,y,k=100,by=type),
        data=prop_geo_pp, bs="cr",
        family=cox.ph(), weight=!active)


pred_surv <- raster(extent(mtl_land),res=100)
crs(pred_surv)<-crs(mtl_land)

# predict survival after 90 days
pred_surv<-raster::interpolate(pred_surv, mmt, fun=my_predict, xyOnly=T,
  const=data.frame(type='Apartment', time_on_market=90 ))

pred_surv <- mask(pred_surv, mask_condo)

m<-tm_shape(mtl_arr,bbox=extent(mtl_arr)) + tm_borders(alpha=0.8, col='black')+
   tm_shape(pred_surv)+tm_raster(style="cont",alpha=0.7, title='p after 90 days')+
   tm_shape(subway_stop_p%>%dplyr::select(stop_name))+tm_symbols(col='blue',alpha=0.2,size=0.03)+
   tm_shape(subway_p%>%dplyr::select(route_name))+tm_lines(col='blue',alpha=0.2)+
   tm_scale_bar(position=c("right", "bottom"))+
   tm_layout(scale=1)

# convert to interactive leaflet map
tmap_leaflet(m)
```

### {data-width=300}
```{r surv_graph}
surv_type<-survfit(Surv(time_on_market, !active) ~ type, prop_geo_pp)

p<-ggsurvplot( surv_type, data=prop_geo_pp, conf.int = TRUE,
    conf.int.style ='step', censor=F, surv.scale='percent',
    break.time.by=30,surv.median.line='hv',
    legend.labs = levels(prop_geo_pp$type),
    xlim=c(0.0,365.0))
p$plot <- p$plot + theme(legend.text = element_text(size = 5, color = "black", face = "bold"))
p
```
